Stage-Splitting a Modal Language

### 2014-09-23 paper/semantics.tex
implemented some suggested changes

The stage-\bbone\ type $\fut A$ contains encapsulated
stage-\bbtwo\ expressions of type $A$. Terms of type $\fut A$ are treated
opaquely by stage-\bbone\ code, as (by the requirements of staging) they cannot be evaluated until stage \bbtwo.
While $\fut$ allows us to embed stage-\bbtwo\ types within stage-\bbone\ types,
there is no way to embed a stage-\bbone\ type within a stage-\bbtwo\ type.
The $\next$ constructor embeds stage-\bbtwo\ expressions into stage \bbone,
while $\prev$ embeds stage-\bbone\ expressions into stage \bbtwo.  $\next$ and $\prev$ are the
only ways in \lang\ to alter the stage of a term; we surround their arguments
with braces in \lang\ syntax to clearly indicate stage boundaries within a
program.

The grammar and type system for \lang\ is given in
\ref{fig:grammar,fig:statics}. (Typing judgments and context variables are annotated with stages.)
Only $\fut$ and its introductory and eliminatory forms $\next$ and $\prev$ affect the stage
of a term or type.
%We formulated our typing judgments in the style of \cite{davies96}, where the
%whole judgment is annotated with a stage.  
%The grammar and type system for \lang\ is given in
%\ref{fig:grammar,fig:statics}.
% We annotate typing judgments and context variables with stages;
%This is made manifest as rules which are entirely abstract over stage.
%In addition to determining the stage, $\next$ and $\prev$ are the introduction and elimination forms for $\fut$ types.
Specifically, given an argument with type $A$ at stage \bbtwo, $\next$ forms a $\fut A$ at stage \bbone.  
%That is, it forms the promise of a future $A$ out of a construction for an $A$ at the next timestep.
Stage \bbtwo\ expressions can obtain the original stage \bbtwo\ argument via the $\prev$ construct.  
Since $\prev$ operates at stage \bbtwo, this ensures no violation of causality\,\cite{cave14}.
The $\pause$ construct serves to wrap stage \bbone\ integers for use in stage \bbtwo.  
It is possible to implement $\pause$ from other \lang\ features, but
we provide it as a core primitive to simplify our examples.

@@

As a simple example of expressing staged programs in \lang, consider the following fast exponentiation algorithm, which 
calculates $b^p$ in $\log p$ time:
\[
	\mathit{fexp}(b,p) = \left \{ \begin{array}{ll} 
		1 &  p = 0 \\ 
		\mathit{fexp}(b,p/2)^2 & p \text{ even} \\ 
		b \cdot \mathit{fexp}(b,p-1) & p \text{ odd} \end{array}
	\right .
\]

% Ignoring for the moment how to define recursive functions in \lang,
\noindent
The stage-\bbone\ term below defines the function {\tt fexp} that implements the fast exponentiation algorithm:

\begin{lstlisting} 
let fexp (b : $int, e : int) : $int =
	if e == 0 then
		next{1}
	else if (e mod 2) == 0 then
		next{let x = prev{fexp(b,p/2)} in x*x}
	else
		next{prev{b} * prev{fexp(b,p-1)}}		
\end{lstlisting}

Although the code for {\tt fexp} appears very similar to the unstaged mathematical
definition above, it is in fact a staged program. The function receives a stage-\bbone\ exponent {\tt e} and a stage-\bbtwo\ base {\tt b} and returns a stage-\bbtwo\ result. The {\tt if} predicates and exponent
decomposition are all stage-\bbone\ terms, since they occur within $\prev$ blocks.
%\needsfix{Essentially, the type system is sufficient to prove the observation that all of
%the decomposition of the exponent can be done at the stage \bbone\ without
%requiring disruption of the elegant functional structure of the code.}
%
% NOTE(kayvonf): removing the following comments about binding time analysis from the text for now to improve clarity.  They will certainly be addressed in some prior section in the paper.
%
%We note that our work does not consider the problem of {\em binding time analysis}---transforming an unstaged program into a staged \lang\ program by inserting appropriate $\next$, $\prev$, and $\pause$ constructs. We assume that binding time analysis is performed manually by a programmer (as done above) or via automatic analysis of unstaged code to arrive at a valid \lang\ program used as input for subsequent splitting.
%
A key attribute of this example is that stage-\bbone\ and stage-\bbtwo\ expressions are nested. Ordinary term evaluation eliminates outermost redexes first, 
however in the case that stage \bbone\ expressions are contained inside
stage \bbtwo\ ones (such as the recurive call to {\tt fexp} above), this strategy conflicts with the precept of staged execution: that all stage-\bbone\ code be evaluated before the evaluation of stage-\bbtwo\ code.

@@

To gain intuition about the challenges of implementing this staged dynamic
semantics, consider the following example:
\begin{lstlisting}
#2 (next {1+2}, 3+4)
\end{lstlisting}
This is a stage-\bbone\ expression of type $\rmint$; the pair inside it is a
stage-\bbone\ expression of type $(\fut\rmint)\times\rmint$. A conventional
call-by-value evaluation strategy demands that we evaluate both components of
the pair to values before we project from the pair. The problem is that
while \verb|next {1+2}| is not a value (in the sense that additional
stage-\bbtwo\ computation steps are necessary to produce the result \verb|3|), evaluating the contents of \verb|next| cannot occur as part of stage-\bbone\ evaluation.
Intuitively, the solution is to designate \verb|next {1+2}| as a value \emph{in
stage \bbone}, even though it requires additional evaluation in stage \bbtwo.
Therefore, we evaluate the pair to
\begin{lstlisting}
#2 (next {1+2}, 7)
\end{lstlisting}
then obtain \verb|7| as the result of projection.

Now consider a more complex example where 
stage-\bbone\ evaluation must substitute such a incompletely-evaluated
expression. The following stage-\bbtwo\ term has type $\rmint$:
\begin{lstlisting} 
prev{
  let x = (next {1+2}, 3+4) in
  next{ prev{#1 x} * prev{#1 x} * hold{#2 x} }
}
\end{lstlisting}
Again, stage-\bbone\ evaluation will not fully reduce this term, because the answer
depends on the value of \verb|1+2|, which is not reduced to \verb|3| until
stage \bbtwo.

If we simply treat \verb|(next {1+2}, 7)| as a value during stage~\bbone, and
substitute it for the three occurrences of \verb|x| in the body of the
\verb|let| expression, the result of stage \bbone\ computation is
\begin{lstlisting} 
prev{
  next{ prev{next {1+2}} * prev{next {1+2}} * 7 }
}
\end{lstlisting}
Finally, the $\prev$s eliminate the $\next$s to yield the final residual:
\begin{lstlisting} 
(1+2) * (1+2) * 7
\end{lstlisting}
Stage-\bbtwo\ evaluation of this expression will compute \verb|1+2| twice. Of
course, this is not particularly expensive, but in general, the duplicated
computation could be anything, even a recursive call to the function! To avoid duplicating
computations, we take a different approach.  Instead of
duplicating the contents of the $\next$ expression in the tuple, we bind the contained stage-\bbtwo\ expression to
a variable (here, $\mathtt{y}$) and duplicate only a reference to that variable.
This produces:
\begin{lstlisting} 
let y = 1+2 in y * y * 7
\end{lstlisting}

Achieving this behavior mechanically requires us to resolve a contradiction:
we must substitute for \texttt{x} in stage \bbone, but we cannot evaluate inside the $\next$ block within the tuple. 
Our solution is to replace the contents of the $\next$ with a new stage-\bbtwo\ variable and create an explicit substitution (shown with a $\mapsto$) binding that stage-\bbtwo\ variable to the $\next$'s old contents.  
This substitution then floats up to the top of the containing $\prev$ block:
\begin{lstlisting} 
prev {
[yhat|->1+2]
  let x = (next{yhat}, 7) in
  next{prev{#1 x} * prev{#1 x} *  hold{#2 x}}
}
\end{lstlisting}
As a convention, we render the new variable with a %stylish and fashionable
hat.  We're now free to perform the stage-\bbone~substitution for {\tt x} without duplicating stage-\bbtwo\ work.
\begin{lstlisting} 
prev {
[y|->1+2]
  next{
    prev{#1 (next {yhat}, 7)} * 
    prev{#1 (next {yhat}, 7)} *
    hold{#2 (next {yhat}, 7)}
  }
}
\end{lstlisting}
To evaluate the outermost $\next$, we must first partially evaluate within its body by finding all of the contained stage-\bbone~terms and reducing them. 
As a rule, these will reduce to $\next$ expressions, which the $\prev$ eliminates, leaving the variable in place:
\begin{lstlisting} 
prev {
[yhat|->1+2]
    next{ yhat * yhat * 7 }
}
\end{lstlisting}
Once again, we lift the contents of the $\next$ into a substitution:
\begin{lstlisting} 
prev {
[yhat|->1+2]
[zhat|->yhat*yhat*7]
    next{ zhat }
}
\end{lstlisting}
Finally, when evaluating the outer $\prev$, we must {\em reify} the contained substitutions into let statements, yielding
\begin{lstlisting} 
let yhat = 1+2 in
let zhat = yhat * yhat * 7 in z
\end{lstlisting}

@@

%{{{ stuff nico commented out
%Previous work (\cite{davies96}) focuses on the correspondence between the type system and existing temporal logics, whereas we care more about operational behavior and cost.  In this section, we'll consider a few proposals for our language before settling on one we like.  All of the proposals are call-by-value, differing primarily in how they handle values of $\fut$ type.

%\subsection{The Erasure Semantics}
%
%We first consider the {\em erasure semantics}, so called because it corresponds to what one would get by interpreting \lamStaged as a single-stage language, ignoring all of the $\next$ and $\prev$ terms.  This gives us two judgments, $\erasone$ and $\erastwo$, corresponding to {\em multistage evaluation} at \bbone and at \bbtwo.  We call these judgments  ``multistage'' because they cause work to happen at both stages.
%
%Both judgments behave normally at non-staged features.  We cover their behavior at staged features below:
%
%\begin{mathpar}
%\infer {\next~e \erasone \next~v} {e \erastwo v} \and
%\infer {\prev~e \erastwo \next~v} {e \erasone v} \and
%\infer {\pause~e \erasone \next~i} {e \erasone i}
%\end{mathpar}
%
%Essentially, we immediately evaluate under the $\next$, yielding a value for $\fut$ types. The $\prev$ terms remove this wrapper.  As expected, $\pause$ also gives us a way to inject into the wrapper at integers.
%
%The erasure semantics has some undesirable properties.  By intention, it interleaves the execution of stage-\bbone~and stage-\bbtwo~code, so the evaluation can't really be said to be staged (i.e. stage-\bbone~work is done before stage-\bbtwo~work).  Moreover, the erasure semantics cannot be equivalent to any semantics which does have this property!  To see why, consider the following code, which types to int at \bbtwo:
%
%\begin{lstlisting} 
%if 5*4*3*2 > 111 then
%	hold{2+4}
%else
%	prev{ loopForever () (* does not terminate *) }
%\end{lstlisting}
%
%Under the erasure semantics (using $\erastwo$), this code takes the top branch and evaluates to 6.  But in order know that the {\tt loopForever} function need not be called, the predicate had to be evaluated prior.  But the predicate is stage-\bbtwo, whereas {\tt loopForever} is stage-\bbone.  To borrow terminology from \cite{cave14}, this violates causality.  In order to avoid this problem, a valid semantics must {\em speculate} down the branches of any stage-\bbtwo~if or case statement (or similarly into the body of a stage-\bbtwo function) to find and evaluate all of the stage-\bbone~code.  Both of the other semantics we will consider have this property.
%
%The benefit of the erasure semantics is that it's very natural, and has a familiar cost model.  It also {\em obviously} produces the ``correct'' answer, so we can use the erasure semantics as a reference to prove the reasonableness of any other semantics.
%
%\subsection{Meta Semantics}
%
%A different semantics was provided in \cite{davies96}.  We briefly review a two-stage version of that semantics here.
%
%In the erasure semantics, $\next v$ is a value only if $v$ is fully reduced.  But in the meta semantics, $\next e$ is a value only if $e$ has no $\prev$ terms; $v$ is allowed to have unreduced stage-2 computation. 
%
%...
%
%The Davies semantics is comprised of two mutually recursive judgments: $\daviesz$ and $\davieso$.  For some $\colone e A$, the $\daviesz$ judgment evaluates only the first-stage parts of $e$, leaving unevaluated second-stage code within $\next$s.  (... This essentially gives us partial evaluation, and we're left to just evaluate the residual normally to get multi-stage evaluation...) 
%
%The benefit of the meta semantics is that it gives us a very explicit notion of partial evaluation.  This also, by construction, means that the meta semantics does all of the first stage-work {\em before} the second stage work begins. 
%
%The $\next$-by-name behavior of the meta semantics of course means that it happily duplicates second-stage code, which could increase the cost.  This makes reasoning about second-stage cost rather difficult.
%
%\begin{lstlisting} 
%let x = next {4+5} in
%next{prev{x} * prev{x}}
%\end{lstlisting}
%
%\subsection{Our Semantics}
%
%We desire a semantics that meets the following goals:
%
%\begin{enumerate}
%\item Modulo termination, it should be equivalent to the erasure semantics.
%\item All of the first stage work should be completed before second stage work.  Ideally, it should just have a notion of partial evaluation, like the meta semantics.
%\item Should be $\next$-by-value, rather than $\next$-by-name, like the erasure semantics.  
%\end{enumerate}
%
%We meet all of these goals.
%}}}
%{{{ stuff carlo commented out
%Abstractly, we can think of our evaluation as proceeding in the standard way for
%stage-1 code. When the evaluator encounters a $\next \{e\}$ expression, it
%places $e$ off to the side in a context and replaces the whole expression with a
%reference to the context entry.  These references are then passed around as
%stage-1 values for $\fut$ types.  But what if $e$ contains $\prev$ expressions?
%To ensure that all stage-1 code is evaluated before any stage-2 code, we must
%evaluate all of the 1-code contained in $e$ before inserting it into the table.
%This entails searching $e$ for all contained $\prev$s and evaluating them in
%place.

%The evaluation judgment operates on stage-1 code, whereas the
%speculation operates on stage-2 code.  Since $\next$ and $\prev$ are the
%crossover points between 1-code and 2-code, they are correspondingly the only
%places where the evaluation and speculation judgments depend on the other. 
%}}}

@@

>>>>

The stage-\bbone\ type $\fut A$ contains encapsulated
stage-\bbtwo\ expressions of type $A$. Terms of type $\fut A$ are treated
opaquely by stage-\bbone\ code, as (by the requirements of staging) they cannot be evaluated until stage \bbtwo.
While $\fut$ allows us to embed stage-\bbtwo\ types within stage-\bbone\ types,
there is no way to embed a stage-\bbone\ type within a stage-\bbtwo\ type.
That is, there is a one-way dependence between stages at the term level.

Alternatively, the stages are mutually dependent at the term level.
The $\next$ constructor embeds stage-\bbtwo\ expressions into stage \bbone,
while $\prev$ embeds stage-\bbone\ expressions into stage \bbtwo.  $\next$ and $\prev$ are the
only ways in \lang\ to alter the stage of a term; we surround their arguments
with braces in \lang\ syntax to clearly indicate stage boundaries within a
program.

The grammar and type system for \lang\ is given in \ref{fig:grammar,fig:statics}. 
Typing judgments and context variables are annotated with stages after an $@$ symbol.
Only $\fut$ and its introductory and eliminatory forms $\next$ and $\prev$ affect the stage
of a term or type.
%We formulated our typing judgments in the style of \cite{davies96}, where the
%whole judgment is annotated with a stage.  
%The grammar and type system for \lang\ is given in
%\ref{fig:grammar,fig:statics}.
% We annotate typing judgments and context variables with stages;
%This is made manifest as rules which are entirely abstract over stage.
%In addition to determining the stage, $\next$ and $\prev$ are the introduction and elimination forms for $\fut$ types.
Specifically, given an argument with type $A$ at stage \bbtwo, $\next$ forms a $\fut A$ at stage \bbone.  
%That is, it forms the promise of a future $A$ out of a construction for an $A$ at the next timestep.
Stage \bbtwo\ expressions can obtain the original stage \bbtwo\ argument via the $\prev$ construct.  
Since $\prev$ operates at stage \bbtwo, this ensures no violation of causality,\cite{cave14}.
The $\pause$ construct serves to wrap stage \bbone\ integers for use in stage \bbtwo.  
It is possible to implement $\pause$ from other \lang\ features, but
we provide it as a core primitive to simplify our examples.

Recall the fast exponent function from above.
We present a valid staging of it here:
\begin{lstlisting} 
let fexp (b : $int, p : int) : $int =
	if p == 0 then
		next{1}
	else if (p mod 2) == 0 then
		next{let x = prev{fexp(b,p/2)} in x*x}
	else
		next{prev{b} * prev{fexp(b,p-1)}}		
\end{lstlisting}

\TODO explain why we staged this the way we did

The function receives a stage-\bbone\ exponent {\tt e} and a stage-\bbtwo\ base {\tt b} and returns a stage-\bbtwo\ result. 
The {\tt if} predicates and exponent decomposition are all stage-\bbone\ terms, since they occur within $\prev$ blocks.

@@

A key attribute of the fexp example is that stage-\bbone\ and stage-\bbtwo\ expressions are nested. 
Ordinary term evaluation eliminates outermost redexes first, 
however in the case that stage \bbone\ expressions are contained inside stage \bbtwo\ ones 
(such as the recurive call to {\tt fexp} above), 
this strategy conflicts with the precept of staged execution: 
that all stage-\bbone\ code be evaluated before the evaluation of stage-\bbtwo\ code.

@@

To gain intuition about the challenges of implementing this staged dynamic
semantics, consider the following example:
\begin{lstlisting}
#2 (next {fib 20}, 2+3)
\end{lstlisting}
This is a stage-\bbone\ expression of type $\rmint$; the pair inside it is a
stage-\bbone\ expression of type $(\fut\rmint)\times\rmint$. 
In this example, ${\tt fib} : \rmint \to \rmint$ is a stage-\bbtwo\ reference to the Fibonacci function.
A conventional call-by-value evaluation strategy demands that we evaluate both components of
the pair to values before we project from the pair. The problem is that
while \verb|next {fib 20}| is not a value (in the sense that additional
stage-\bbtwo\ computation steps are necessary to produce the result \verb|6765|), 
evaluating the contents of \verb|next| cannot occur as part of stage-\bbone\ evaluation.
Intuitively, the solution is to designate \verb|next {fib 20}| as a value \emph{in
stage \bbone}, even though it requires additional evaluation in stage \bbtwo.
Therefore, we evaluate the pair to
\begin{lstlisting}
#2 (next {fib 20}, 5)
\end{lstlisting}
then obtain \verb|5| as the result of projection.

Now consider a more complex example where 
stage-\bbone\ evaluation must substitute such a incompletely-evaluated
expression. The following stage-\bbtwo\ term has type $\rmint$:
\begin{lstlisting} 
prev{
  let x = (next {fib 20}, 3+4) in
  next{ prev{#1 x} * prev{#1 x} * hold{#2 x} }
}
\end{lstlisting}
Again, stage-\bbone\ evaluation will not fully reduce this term, because the answer
depends on the value of \verb|fib 20|, which is not reduced to \verb|3| until
stage \bbtwo.

If we simply treat \verb|(next {fib 20}, 7)| as a value during stage~\bbone, and
substitute it for the three occurrences of \verb|x| in the body of the
\verb|let| expression, the result of stage \bbone\ computation is
\begin{lstlisting} 
prev{
  next{ 
    prev{next {fib 20}} * prev{next {fib 20}} * 7 
  }
}
\end{lstlisting}
Finally, the $\prev$s eliminate the $\next$s to yield the final residual:
\begin{lstlisting} 
(fib 20) * (fib 20) * 7
\end{lstlisting}
Note how stage-\bbtwo\ evaluation of this expression will compute \verb|fib 20| twice.  
To avoid duplicating computations, we take a different approach.  Instead of
duplicating the contents of the $\next$ expression in the tuple, we bind the contained stage-\bbtwo\ expression to
a variable (here, $\mathtt{y}$) and duplicate only a reference to that variable.
This produces:
\begin{lstlisting} 
let y = fib 20 in y * y * 7
\end{lstlisting}

Achieving this behavior mechanically requires us to resolve a contradiction:
we must substitute for \texttt{x} in stage \bbone, but we cannot evaluate inside the $\next$ block within the tuple. 
Our solution is to replace the contents of the $\next$ with a new stage-\bbtwo\ variable and create an explicit substitution (shown with a $\mapsto$) binding that stage-\bbtwo\ variable to the $\next$'s old contents.  
This substitution then floats up to the top of the containing $\prev$ block:
\begin{lstlisting} 
prev {
[yhat|->fib 20]
  let x = (next{yhat}, 7) in
  next{prev{#1 x} * prev{#1 x} *  hold{#2 x}}
}
\end{lstlisting}
As a convention, we render the new variable with a %stylish and fashionable
hat.  We're now free to perform the stage-\bbone~substitution for {\tt x} without duplicating stage-\bbtwo\ work.
\begin{lstlisting} 
prev {
[yhat|->fib 20]
  next{
    prev{#1 (next {yhat}, 7)} * 
    prev{#1 (next {yhat}, 7)} *
    hold{#2 (next {yhat}, 7)}
  }
}
\end{lstlisting}
To evaluate the outermost $\next$, we must first partially evaluate within its body by finding all of the contained stage-\bbone~terms and reducing them. 
As a rule, these will reduce to $\next$ expressions, which the $\prev$ eliminates, leaving the variable in place:
\begin{lstlisting} 
prev {
[yhat|->fib 20]
    next{ yhat * yhat * 7 }
}
\end{lstlisting}
Once again, we lift the contents of the $\next$ into a substitution:
\begin{lstlisting} 
prev {
[yhat|->fib 20]
[zhat|->yhat*yhat*7]
    next{ zhat }
}
\end{lstlisting}
Finally, when evaluating the outer $\prev$, we must {\em reify} the contained substitutions into let statements, yielding
\begin{lstlisting} 
let yhat = fib 20 in
let zhat = yhat * yhat * 7 in z
\end{lstlisting}

@@

@@

Returning to our {\tt fexp} example, we can speculate on the stage \bbtwo\ term
$\verb|fn b => prev{fexp(next{b},10)}|$
to get the residual
\begin{lstlisting} 
fn b : int =>
  let x0 =
    (let x1 = 
      let x2 = 1 * b
      in x2 * x2
    in x1 * x1) * b
  in x0 * x0
\end{lstlisting}

Note how the recursion causes duplicate code in the residual.

### 2014-09-23 paper/techreport.tex
implemented some suggested changes

With these goals in mind, our initial concern is how to express multistage programs. To this end, we have modified prior work (\cite{davies96}) to create \lang, a typed multi-stage lambda calculus featuring products, sums, and isorecursive types. An important property of \lang\ is that the staging of a well-typed program is apparent without resorting to any variable analysis, contra standard binding time analysis.

@@

>>>>

With these goals in mind, our initial concern is how to express multistage programs. To this end, we have created \lang, a typed multi-stage lambda calculus featuring products, sums, and isorecursive types. An important property of \lang\ is that the staging of a well-typed program is apparent without resorting to any variable analysis, contra standard binding time analysis.

@@

\TODO motivate this example
\begin{lstlisting} 
let fexp (b : int, p : int) : int =
	if p == 0 then
		1
	else if (p mod 2) == 0 then
		let x = fexp(b,p/2) in x*x
	else
		b * fexp(b,p-1)	
\end{lstlisting}
